---
title: "Tidymodels"
output: 
    html_notebook:
        number_sections: true
        toc: true
        toc_float: true
        toc_depth: 3
        toc_collapsed: true
---

# Workflow with an example

# Learn tidymodels

Following this [tutorial](https://www.tidymodels.org/start/models/).

## Build a model

### Prepare data

```{r}
library(tidymodels)  # for the parsnip package, along with the rest of tidymodels

# Helper packages
library(readr)       # for importing data
library(broom.mixed) # for converting bayesian models to tidy tibbles
library(dotwhisker)  # for visualizing regression results

# data
urchins <-
  # Data were assembled for a tutorial 
  # at https://www.flutterbys.com.au/stats/tut/tut7.5a.html
  read_csv("https://tidymodels.org/start/models/urchins.csv") %>% 
  # Change the names to be a little more verbose
  setNames(c("food_regime", "initial_volume", "width")) %>% 
  # Factors are very helpful for modeling, so we convert one column
  mutate(food_regime = factor(food_regime, levels = c("Initial", "Low", "High")))

head(urchins)

# visualize the data
ggplot(urchins,
       aes(x = initial_volume, 
           y = width, 
           group = food_regime, 
           col = food_regime)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE) +
  scale_color_viridis_d(option = "plasma", end = .7)
```


### Build and fit model
The response variable is `width`. Let's first build a linear regression model.

```{r}
# define the model
lm_mod <- 
  linear_reg() %>%   # type of model
  set_engine("lm")   # method to fitting / training

# fit the model
lm_fit <- 
  lm_mod %>%
  # specify data and formula, considering interaction term
  fit(width ~ initial_volume * food_regime, data = urchins)

# view model parameters
tidy(lm_fit)

# visualize the model parameters
tidy(lm_fit) %>%
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, color = "gray50", linetype = 2))
```
### Make prediction

```{r}
# make up new data
new_points <- expand.grid(initial_volume = 20, 
                          food_regime = c("Initial", "Low", "High"))

# make prediction
mean_pred <- predict(lm_fit, new_data = new_points)
mean_pred

conf_int_pred <- predict(lm_fit, new_data = new_points, type = "conf_int")
conf_int_pred

# combine mean and confidence intervals
pred_data <- 
  new_points %>%
  bind_cols(mean_pred) %>%
  bind_cols(conf_int_pred)
pred_data

# visualize predictions
pred_data %>%
  ggplot(aes(x = food_regime)) +
  geom_point(aes(y = .pred)) +
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = 0.2) +
  labs(y = "urchine size")
```

### Change model engine

Try a Bayesian analysis.

```{r}
# fit bayes model
bayes_fit <- 
  linear_reg() %>%
  # raplace engine here
  set_engine("stan",
             prior_intercept = rstanarm::student_t(df = 1),
             prior = rstanarm::student_t(df = 1)) %>%
  fit(width ~ initial_volume * food_regime, data = urchins)
print(bayes_fit, digits = 5)
tidy(bayes_fit, conf.int = TRUE)

# make predictions
new_points %>%
  bind_cols(predict(bayes_fit, new_data = new_points)) %>%
  bind_cols(predict(bayes_fit, new_data = new_points, type = "conf_int")) %>%
  ggplot(aes(x = food_regime)) +
  geom_point(aes(y = .pred)) +
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = 0.2) +
  labs(y = "urchine size")
```

## Preprocess data with recipes

### Prepare data

```{r}
library(tidymodels)      # for the recipes package, along with the rest of tidymodels

# Helper packages
library(nycflights13)    # for flight data
library(skimr)           # for variable summaries

set.seed(123)

flight_data <- 
  flights %>% 
  mutate(
    # Convert the arrival delay to a factor
    arr_delay = ifelse(arr_delay >= 30, "late", "on_time"),
    arr_delay = factor(arr_delay),
    # We will use the date (not date-time) in the recipe below
    date = as.Date(time_hour)
  ) %>% 
  # Include the weather data
  inner_join(weather, by = c("origin", "time_hour")) %>% 
  # Only retain the specific columns we will use
  select(dep_time, flight, origin, dest, air_time, distance, 
         carrier, date, arr_delay, time_hour) %>% 
  # Exclude missing data
  na.omit() %>% 
  # For creating models, it is better to have qualitative columns
  # encoded as factors (instead of character strings)
  mutate_if(is.character, as.factor)

# 16% arrive late
flight_data %>%
  count(arr_delay) %>%
  mutate(prop = n / sum(n))

# make sure the response variable arr_delay is a factor
glimpse(flight_data)

# check factors
flight_data %>%
  skimr::skim(dest, carrier)
```

### train-test splitting

```{r}
set.seed(555)
data_split <- initial_split(flight_data, prop = 3/4)
train_data <- training(data_split)
test_data <- testing(data_split)
```

### Create recipe, roles, and features

```{r}
flights_rec <- 
  # initiate a recipe
  recipe(arr_delay ~ ., data = train_data) %>%
  # flight and time_hour are not used for modeling
  update_role(flight, time_hour, new_role = "ID") %>%
  # features derived from date
  step_date(date, features = c("dow", "month")) %>%   # dow: day of week
  step_holiday(date, holidays = timeDate::listHolidays("US")) %>%
  step_rm(date) %>%   # remove date
  # covert categorical varaibles into dummy variables for logistic regression
  step_dummy(all_nominal(), -all_outcomes()) %>%
  # remove zero-variance variables, often for dummy variables that are rare
  step_zv(all_predictors())


# examine the recipe
summary(flights_rec)
```

### Fit model with recipe

```{r}
# setup logistic regression model
lr_mod <- 
  logistic_reg() %>% 
  set_engine("glm")

# create workflow
flights_wflow <- 
  workflow() %>% 
  add_model(lr_mod) %>% 
  add_recipe(flights_rec)
flights_wflow

# fit model
flights_fit <-
  flights_wflow %>%
  fit(data = train_data)
      
# view model parameters
flights_fit %>%
  extract_fit_parsnip() %>%
  tidy()
```

### Make prediction on test data

```{r}
# predict the outcome
predict(flights_fit, test_data)

# predict probability
flights_pred <-
  predict(flights_fit, test_data, type = "prob") %>%
  bind_cols(test_data %>% select(arr_delay, time_hour, flight))
flights_pred

# ROC curve
flights_pred %>%
  roc_curve(truth = arr_delay, .pred_late) %>%
  autoplot()

# ROC AUC score
flights_pred %>%
  roc_auc(truth = arr_delay, .pred_late)
```


