---
title: "Use MySQL in R"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Prepare MySQL for R
Following instructions [here](https://github.com/r-dbi/RMySQL). But always search for latest instructions.

### Install drivers for Ubuntu
`UnixODBC` library is required for all databases
```
$ apt-get install unixodbc unixodbc-dev --install-suggests
```

Then install specific for MySQL
```
$ sudo apt-get install -y libmariadbclient-dev
```


### Set up database connections
Here is a [good summary](https://www.r-bloggers.com/2020/09/how-to-connect-r-with-sql/) on how to make connections.


## Build a sql database of geography using 2010 decennial census

```{r}
library(data.table)
library(magrittr)
library(DBI)
library(totalcensus)
library(stringr)

# connect to database in MySQL server
con <- dbConnect(RMySQL::MySQL(), 
                 dbname = "geography", 
                 host = "localhost",    # remote host like "https://abc123.com/"
                 # will ask for usename and password in popup box
                 # user = rstudioapi::askForPassword("User name"), 
                 user = "gl",
                 password = rstudioapi::askForPassword("Password"))
# list tables
dbListTables(con)
# list field in table rental
dbListFields(con, "County")
```

## insert values from R
To write a dataframe to a database table, the dataframe must satisfy these requirements:

- Column names of the dataframe must be a subset of those of the database table.
- The columns of database table that have no default values or auto_increment must present in the dataframe.
- No need to be in the same order.

### State table
```{r}
# get the data
state <- read_decennial(2010, "US", summary_level = "state")
state_sql <- state[, .(fips = str_sub(GEOID, 8, 9),
                       name = NAME,
                       abbr = state,
                       population, lon, lat,
                       geoid = GEOID,
                       geoid_short = str_extract(GEOID, "[0-9]+$"))]

# in case to reset the primary to start from 1, from mysql command line
# ALTER TABLE State AUTO_INCREMENT=1;
dbWriteTable(con, "State", state_sql, append = TRUE)

# check the table from mysql command line and get id and abbr for later use
state_db <- dbReadTable(con, "State") %>%
    setDT() %>%
    .[, .(state_id, state = abbr)]
```

### Metro table
```{r}
metro <- read_decennial(2010, "US", summary_level = "310")
metro_sql <- metro[, .(fips = str_sub(GEOID, 8, 12),
                       name = NAME, 
                       population, lon, lat,
                       geoid = GEOID,
                       geoid_short = str_extract(GEOID, "[0-9]+$"))]

dbWriteTable(con, "Metro", metro_sql, append = TRUE)

metro_db <- dbReadTable(con, "Metro") %>%
    setDT() %>%
    .[, .(metro_id, metro_fips = fips)]
```


### Place table
```{r}
# get data. state is still in abbr not in state_id
place <- read_decennial(2010, "US", summary_level = "place")
place_sql <- place[, .(fips = str_sub(GEOID, 10, 14),
                       name = str_remove(NAME, ",[^,]+$"),
                       state, population, lon, lat,
                       geoid = GEOID,
                       geoid_short = str_extract(GEOID, "[0-9]+$"))] %>%
    # convert state to state_id
    state_db[., on = .(state)] %>%
    # reorder column to match Place table
    .[, .(fips, name, state_id, population, lon, lat, geoid, geoid_short)]

# write to table
dbWriteTable(con, "Place", place_sql, append=TRUE)
```


### County table
```{r}
county <- read_decennial(2010, "US", geo_headers = "CBSA", 
                        summary_level = "county")
county_sql <- county[, .(fips = str_sub(GEOID, 10, 12),
                       name = NAME,
                       state, 
                       metro_fips = CBSA,
                       population, lon, lat,
                       geoid = GEOID,
                       geoid_short = str_extract(GEOID, "[0-9]+$"))] %>%
    state_db[., on = .(state)] %>%
    metro_db[., on = .(metro_fips)] %>%
    .[, .(fips, name, state_id, metro_id, population, lon, lat, geoid, geoid_short)]

dbWriteTable(con, "County", county_sql, append = TRUE)

county_db <- dbReadTable(con, "County") %>%
    setDT() %>%
    .[, .(county_id, county_geoid_short = geoid_short)]
```


### County_Subdivision table

```{r}
cousub <- read_decennial(2010, "US", geo_headers = "CBSA", 
                        summary_level = "060")

cousub_sql <- cousub[, .(fips = str_sub(GEOID, 12, 16),
                         name = NAME,
                         population, lon, lat, state,
                       geoid = GEOID,
                       geoid_short = str_extract(GEOID, "[0-9]+$"),
                       metro_fips = CBSA)] %>%
    .[, county_geoid_short := str_sub(geoid_short, 1, 5)] %>%
    state_db[., on = .(state)] %>%
    metro_db[., on = .(metro_fips)] %>%
    county_db[., on = .(county_geoid_short)] %>%
    .[, .(fips, name, population, state_id, county_id, metro_id, lon, lat,
          geoid, geoid_short)]
    
dbWriteTable(con, "County_Subdivision", cousub_sql, append = TRUE)
```



### disconnect from a database
```{r}
dbDisconnect(con)
```
