---
title: "Multilevel Regression"
author: "GL"
date: "7/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Load packages and data

```{r}
library(lme4)
library(lmerTest)
library(tidyverse)
library(data.table)
library(ggplot2)

dat <- fread("SCI.DAT") %>%
    setnames(c("center_id", "candidate_id", "gender",
               "score_paper", "score_teacher")) %>%
    # center contineous independent variables
    .[, score_paper := score_paper - mean(score_paper)]
```



## The simplest mixed model - interception only

The formula can be `score_teacher ~ 1 + (1 | center_id)`. They have the same outcome.

```{r}
m1 <- lmer(score_teacher ~ (1 | center_id), data = dat)
```

### Understand randomf and fixed effect

```{r}
summary(m1)
```


### Understand the prediction

The prediction is approximately the mean of each center_id but biased toward the global mean. The fewer the samples, the more the bias. The center_id 84707 has the largest bias, which only has two data points. The center_id 68207 has the second largest bias, which has 5 data points.

```{r}
# predicted value for each center_id
pred <- predict(m1, dat)
df <- data.table(center_id = dat$center_id, pred) %>% 
    unique()

# mean of each center_id
avg <- dat[, .(avg = mean(score_teacher),
               n = .N), by = .(center_id)]

# combine together
df_compare <- df[avg, on = "center_id"] %>%
    .[, difference := pred - avg]
ggplot(df_compare, aes(avg, pred, size = n)) +
    geom_point(alpha = 0.5) +
    scale_size_area(breaks = c(2, 10, 50, 100)) +
    geom_abline(slope = 1, intercept = 0) +
    geom_hline(yintercept = mean(dat$score_teacher), linetype = "dotted") +
    labs(title = "Compare True Average and Prediciton of Each Center",
         x = "True Average",
         y = "Prediction",
         size = "Number of\nSamples") +
    theme_bw()
```



## Treat gender as level 2 unit
```{r}
m2 <- glmer(score_teacher ~ (1 | center_id) + (1 | gender), data = dat)

# predicted value for each center_id
pred <- predict(m2, dat)
dat$pred <- pred
df <- data.table(center_id = dat$center_id, 
                 gender = dat$gender,
                 pred = pred) %>% 
    .[, .(pred = unique(pred)), by = .(center_id, gender)]

# mean of each center_id
avg <- dat[, .(avg = mean(score_teacher),
               n = .N),
           by = .(center_id, gender)]

# combine together
df_compare <- df[avg, on = c("center_id", "gender")] %>%
    .[, difference := pred - avg]
ggplot(df_compare, aes(avg, pred, size = n)) +
    geom_point(alpha = 0.5) +
    scale_size_area(breaks = c(1, 10, 50, 100)) +
    geom_abline(slope = 1, intercept = 0) +
    geom_hline(yintercept = mean(dat$score_teacher), linetype = "dotted") +
    facet_wrap(~gender) +
    labs(title = "Compare True Average and Prediciton of Each Center and Gender",
         x = "True Average",
         y = "Prediction",
         size = "Number of\nSamples") +
    theme_bw()
```

```{r}
ggplot(df_compare, aes(n, difference, color = (pred > mean(dat$score_teacher)))) +
    geom_point()
```




## Treat gender as level 1 variable
The model outcome is very similar to m2 where gender is treated as a level 2 unit.

```{r}
m3 <- lmer(score_teacher ~ gender + (1 | center_id), data = dat)
```

### Understand random and fixed effects
```{r}
summary(m3)
```

### Understand prediction

```{r}
pred <- predict(m3, dat)
dat$pred <- pred
 
df <- data.table(center_id = dat$center_id, 
                 gender = dat$gender,
                 pred = pred) %>% 
    .[, .(pred = unique(pred)), by = .(center_id, gender)]

# mean of each center_id
avg <- dat[, .(avg = mean(score_teacher),
               n = .N),
           by = .(center_id, gender)]

# combine together
df_compare <- df[avg, on = c("center_id", "gender")] %>%
    .[, difference := pred - avg]
ggplot(df_compare, aes(avg, pred, size = n)) +
    geom_point(alpha = 0.5) +
    scale_size_area(breaks = c(1, 10, 50, 100)) +
    geom_abline(slope = 1, intercept = 0) +
    geom_hline(yintercept = mean(dat$score_teacher), linetype = "dotted") +
    facet_wrap(~gender) +
    labs(title = "Compare True Average and Prediciton of Each Center and Gender",
         x = "True Average",
         y = "Prediction",
         size = "Number of\nSamples") +
    theme_bw()
```


# Interaction between center_id and gender

## Treat gender as level 2 unit
```{r}
m4 <- glmer(score_teacher ~ (1 | center_id : gender), data = dat)

# predicted value for each center_id
pred <- predict(m4, dat)
dat$pred <- pred
df <- data.table(center_id = dat$center_id, 
                 gender = dat$gender,
                 pred = pred) %>% 
    .[, .(pred = unique(pred)), by = .(center_id, gender)]

# mean of each center_id
avg <- dat[, .(avg = mean(score_teacher),
               n = .N),
           by = .(center_id, gender)]

# combine together
df_compare <- df[avg, on = c("center_id", "gender")] %>%
    .[, difference := pred - avg]
ggplot(df_compare, aes(avg, pred, size = n)) +
    geom_point(alpha = 0.5) +
    scale_size_area(breaks = c(1, 10, 50, 100)) +
    geom_abline(slope = 1, intercept = 0) +
    geom_hline(yintercept = mean(dat$score_teacher), linetype = "dotted") +
    # facet_wrap(~gender) +
    labs(title = "Compare True Average and Prediciton of Each Center and Gender",
         x = "True Average",
         y = "Prediction",
         size = "Number of\nSamples") +
    theme_bw()
```