---
title: "Multilevel Regression"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```


## Introduction

### Why multilevel regression
Generalized linear models assume that observations are independent from each other, which often is not true. For example, the people from the same county will have closer attributes such as income and education levels than in other countries in a world population study. Multilevel regression addresses this issue by considering the difference at country level.

### Terminologies

In a study we want to study the relationship between school scores and IQ of students in a school district. The school district has 10 schools of various size. The collected data has the variables school_id, student_id, school_size, student_score, and student_iq. 

- school_score is the dependent variable
- student_id is the id of level 1 unit
- school_id is the id of level 2 unit
- student_iq is a level 1 variable
- school_size is a level 2 variable


## Load packages and data

```{r}
library(lme4)
library(lmerTest)
library(tidyverse)
library(data.table)
library(ggplot2)

dat <- fread("SCI.DAT") %>%
    setnames(c("center_id", "candidate_id", "gender",
               "score_paper", "score_teacher")) %>%
    # center contineous independent variables
    .[, score_paper := score_paper - mean(score_paper)]

head(dat)
```
In the table

- `score_teacher` is the dependent variables
- `candidate_id` and `score_paper` are level 1 predictors and
- `center_id` are a level 2 predictor.

Let's only consider one level 1 predictor, `score_paper`, the `score_teacher` of the $i^{th}$ candidate in the $j^{th}$ center is 

$$\mbox{score_teacher}_{ij} = \beta_{0j} + \beta_{1j} \times \mbox{score_paper} + \epsilon_{ij}$$
Where $\beta_{0j}$ and $\beta_{1j}$ varies from cneter to center and $\epsilon_{ij}$ is the error at level 1, which is associated with fix effect (discussed later).

Since $\beta_{0j}$ and $\beta_{1j}$ depend on center, we can write them in the form of

$$\beta_{0j} = \gamma_{00} + \gamma_{01} \times \mbox{center_id} + u_{0j}$$
$$\beta_{1j} = \gamma_{10} + \gamma_{11} \times \mbox{center_id} + u_{1j}$$
where $\beta_{0j}$ and $\beta_{1j}$ are considered as dependent varaibles and level 2 variable `center_id` is considered as predictor. 

## The simplest mixed model - interception only

The formula can be `score_teacher ~ 1 + (1 | center_id)`. They have the same outcome.

```{r}
m1 <- lmer(score_teacher ~ (1 | center_id), data = dat)
```

### Understand random and fixed effect

```{r}
summary(m1)
```


### Understand the prediction

The prediction is approximately the mean of each center_id but biased toward the global mean. The fewer the samples, the more the bias. The center_id 84707 has the largest bias, which only has two data points. The center_id 68207 has the second largest bias, which has 5 data points.

```{r}
# predicted value for each center_id
pred <- predict(m1, dat)
df <- data.table(center_id = dat$center_id, pred) %>% 
    unique()

# mean of each center_id
avg <- dat[, .(avg = mean(score_teacher),
               n = .N),
           by = .(center_id)]

# combine together
df_compare <- df[avg, on = "center_id"] %>%
    .[, difference := pred - avg]
ggplot(df_compare, aes(avg, pred, size = n)) +
    geom_point(alpha = 0.5) +
    scale_size_area(breaks = c(2, 10, 50, 100)) +
    geom_abline(slope = 1, intercept = 0) +
    geom_hline(yintercept = mean(dat$score_teacher), linetype = "dotted") +
    labs(title = "Compare True Average and Prediciton of Each Center",
         x = "True Average",
         y = "Prediction",
         size = "Number of\nSamples") +
    theme_bw()
```



## Treat gender as level 2 unit
```{r}
m2 <- glmer(score_teacher ~ (1 | center_id) + (1 | gender), data = dat)

# predicted value for each center_id
pred <- predict(m2, dat)
dat$pred <- pred
df <- data.table(center_id = dat$center_id, 
                 gender = dat$gender,
                 pred = pred) %>% 
    .[, .(pred = unique(pred)), by = .(center_id, gender)]

# mean of each center_id
avg <- dat[, .(avg = mean(score_teacher),
               n = .N),
           by = .(center_id, gender)]

# combine together
df_compare <- df[avg, on = c("center_id", "gender")] %>%
    .[, difference := pred - avg]
ggplot(df_compare, aes(avg, pred, size = n)) +
    geom_point(alpha = 0.5) +
    scale_size_area(breaks = c(1, 10, 50, 100)) +
    geom_abline(slope = 1, intercept = 0) +
    geom_hline(yintercept = mean(dat$score_teacher), linetype = "dotted") +
    facet_wrap(~gender) +
    labs(title = "Compare True Average and Prediciton of Each Center and Gender",
         x = "True Average",
         y = "Prediction",
         size = "Number of\nSamples") +
    theme_bw()
```

```{r}
ggplot(df_compare, aes(n, difference, color = (pred > mean(dat$score_teacher)))) +
    geom_point()
```




## Treat gender as level 1 variable
The model outcome is very similar to m2 where gender is treated as a level 2 unit.

```{r}
m3 <- lmer(score_teacher ~ gender + (1 | center_id), data = dat)
```

### Understand random and fixed effects
```{r}
summary(m3)
```

### Understand prediction

```{r}
pred <- predict(m3, dat)
dat$pred <- pred
 
df <- data.table(center_id = dat$center_id, 
                 gender = dat$gender,
                 pred = pred) %>% 
    .[, .(pred = unique(pred)), by = .(center_id, gender)]

# mean of each center_id
avg <- dat[, .(avg = mean(score_teacher),
               n = .N),
           by = .(center_id, gender)]

# combine together
df_compare <- df[avg, on = c("center_id", "gender")] %>%
    .[, difference := pred - avg]
ggplot(df_compare, aes(avg, pred, size = n)) +
    geom_point(alpha = 0.5) +
    scale_size_area(breaks = c(1, 10, 50, 100)) +
    geom_abline(slope = 1, intercept = 0) +
    geom_hline(yintercept = mean(dat$score_teacher), linetype = "dotted") +
    facet_wrap(~gender) +
    labs(title = "Compare True Average and Prediciton of Each Center and Gender",
         x = "True Average",
         y = "Prediction",
         size = "Number of\nSamples") +
    theme_bw()
```


## Interaction between center_id and gender

## Treat gender as level 2 unit
```{r}
m4 <- glmer(score_teacher ~ (1 | center_id : gender), data = dat)

# predicted value for each center_id
pred <- predict(m4, dat)
dat$pred <- pred
df <- data.table(center_id = dat$center_id, 
                 gender = dat$gender,
                 pred = pred) %>% 
    .[, .(pred = unique(pred)), by = .(center_id, gender)]

# mean of each center_id
avg <- dat[, .(avg = mean(score_teacher),
               n = .N),
           by = .(center_id, gender)]

# combine together
df_compare <- df[avg, on = c("center_id", "gender")] %>%
    .[, difference := pred - avg]
ggplot(df_compare, aes(avg, pred, size = n)) +
    geom_point(alpha = 0.5) +
    scale_size_area(breaks = c(1, 10, 50, 100)) +
    geom_abline(slope = 1, intercept = 0) +
    geom_hline(yintercept = mean(dat$score_teacher), linetype = "dotted") +
    # facet_wrap(~gender) +
    labs(title = "Compare True Average and Prediciton by Center and Gender",
         x = "True Average",
         y = "Prediction",
         size = "Number of\nSamples") +
    theme_bw()
```