---
title: "Application of statistics in real projects"
output: 
    html_notebook:
        toc: true
        number_section: false
---


# Project 1: NHANES data analysis

## Take-home messages

### Understand different weights

- **survey weight** is the number of people each respondent represents.  It is different from those in machine learning algorithms.
- **machine learning weight** specifies the importance of a class or case when calculating lost function.


### Use survey package for survey data analysis

Survey data analysis has its own ecosystem in R.


### Use weights package for quick analysis of weighted data
- `wtd.t.test`
- `wtd.chi.sq`
- `wtd.cor`
- `wtd.hist`




## Data and libraries

```{r}
# example data: all NHAA population in 2011 - 2018 NHANES data
library(tidyverse)
library(survey)
library(weights)
nhanes <- readRDS("asian_final.RDS")
```


## Report one sample mean / proportion with 95% confidence interval

- mean age of one group
- mean proportion of female in one group

### Mean of one weighted group with 95% confidence interval

**Case study**: Age of participants was collected in a survey. To describe the age distribution, it is a standard and the most succinct way to report the mean of age with 95% confidence interval, in a form like 37.1 (35.5, 38.7). 

```{r}
report_mean <- function(x, wt = NULL, 
                        conf_level = 0.95, 
                        output = "string", 
                        digits = 3){
    # Report the mean and confidence interval of numeric vector of one group 
    # with weight. It is the result of one sample t-test, comparing 
    # mean against zero.
    # 
    # Arguments:
    # - x: numeric vector
    # - wt: numeric vector of weight, same length as x
    # - conf_level: numeric, confidence level
    # - output: string, output format. If "string", the output is a string like
    #   "3.14 (2.98, 3.30)". If "list", the output is a list of element "avg", 
    #   "upper", "lower".
    # - digits: integer, number of significant digits to keep in the "string" 
    #   outupt.
    #
    # Return: a string or a list depending
    
    df <- length(x) - 1  # degree of freedom
    
    if (is.null(wt)){
        wt <- rep(1, length(x))
    }
    
    wtd_t_test <- weights::wtd.t.test(x = x, weight = wt)
    avg <- wtd_t_test$additional[2]
    std_err <- wtd_t_test$additional[4]
    t_stat <- avg/std_err ## not mx - mu, how differ from zero
    alpha <- 1 - conf_level
    conf_int <- qt(1 - alpha/2, df)
    conf_int <- t_stat + c(-conf_int, conf_int)
    ci <- conf_int * std_err
    
    if (output == "string"){
        avg <- signif(avg, digits)
        low <- signif(ci[1], digits)
        up <- signif(ci[2], digits)
        cat(paste0(avg, " (", low, ", ", up, ")"))
        return(invisible(NULL))
    } else if (output == "list"){
        return(list(avg = avg, lower = ci[1], upper = ci[2]))
    }
}

# example
report_mean(x = nhanes$age,
            wt = nhanes$weight_interview)
```

### Report proportion of one sample

**Case study**: Gender was collect in a survey. To describe the proportion of female, we can report the proportion with 95% confidence interval, like 0.45 (0.41, 0.49).

Confidence interval of proportion cannot be calculated with t-test. To include weight, we will use [survey based confidence interval](https://stackoverflow.com/questions/60636060/how-to-determine-the-confidence-interval-for-a-weighted-population-proportion-wi). Its result, however, approaches that of t-test when sample size is large.

```{r}
report_proportion <- function(x, wt = 1, 
                              conf_level = 0.95, 
                              output = "string", 
                              digits = 3){
    # Report the proportion and confidence interval of being TRUE of a logic
    # vector with weight. 
    # 
    # Arguments:
    # - x: logic vector of TRUE and FALSE
    # - wt: numeric vector of weight, same length as x
    # - conf_level: numeric, confidence level
    # - output: string, output format. If "string", the output is a string like
    #   "3.14 (2.98, 3.30)". If "list", the output is a list of element "avg", 
    #   "upper", "lower".
    # - digits: integer, number of significant digits to keep in the outupt.
    #
    # Return: a string or a list depending
    
    if (length(wt) == 1){
        wt <- rep(wt, length(x))
    }
    
    df <- data.frame(x = x, wt = wt) 
    
    svy_design <- survey::svydesign(ids = ~1, weights = wt, data = df)
    design_mean <- survey::svymean(~x, svy_design, deff="replace")
    ci <- confint(design_mean, level = conf_level)
    
    if (output == "string"){
        avg <- signif(design_mean[2], digits) * 100  # values for TRUE
        low <- signif(ci[2, 1], digits) * 100  
        up <- signif(ci[2, 2], digits) * 100
        cat(paste0(avg, " (", low, ", ", up, ")"))
        return(invisible(NULL))
    } else if (output == "list"){
        return(list(avg = design_mean[2], 
                    lower = ci[2, 1],     
                    upper = ci[2, 2]))
    }
}

# example
x = nhanes$gender
wt = nhanes$weight_interview
report_proportion(x == "F", wt)
```

## Compare two sample means / proportions

We will compare

- mean age of two groups and p-value of t-test
- mean proportion of female of two groups and p-value of Chi-square test

and the final out put is a table in the format

characteristics | reported screening | not reported screening | p value
:-----          | :----              | :----                  | :----
N sample        | 891                | 646                    |
Age (years)     | 40.5 (39.5, 41.5)  | 36.4 (35.6, 37.2)      | < 0.00001
Female (%)      | 36.2 (32.2, 40.2)  | 37.8 (34.5, 41.2)      | 0.51

### Compare two-sample means

**Case study**: To describe the participant's age in groups who reported and not reported diabetes screening. We can report the mean with 95% confidence interval of each group and the p-value of t-test of the two group mean.



```{r}
# use library weights for weighted t test

compare_two_mean <- function(x1, x2, wt1, wt2){
    # To report the mean (95% confidence interval) of two groups and the p-value
    # of two sample t-test
    #
    # Argument: see those in function report_mean
    
    # print mean and confidence interval of each group
    report_mean(x1, wt1)
    cat("\n")
    
    report_mean(x2, wt2)
    cat("\n")
    
    # report p-value of two sample t-test with weight
    tt <- weights::wtd.t.test(x1, x2, wt1, wt2)
    print(tt)
}

# example
x1 <- nhanes %>% filter(screened == 1) %>% pull(age)
wt1 <- nhanes %>% filter(screened == 1) %>% pull(weight_interview)
x2 <- nhanes %>% filter(screened == 0) %>% pull(age)
wt2 <- nhanes %>% filter(screened == 0) %>% pull(weight_interview)
compare_two_mean(x1, x2, wt1, wt2)
```

### Compare two sample proportions

**Case study**: To describe the participant's gender in groups who reported and not reported diabetes screening. We can report the proportion with 95% confidence interval of each group and the p-value of chi-square test of the two group proportion.


```{r}
compare_two_proportion <- function(x1, x2, wt1, wt2){
    # To report the proportion (95% confidence interval) of two groups and the 
    # p-value of Chi-squared test
    #
    # Argument: see those in function report_proportion
    
    # print mean and confidence interval of each group
    report_proportion(x1, wt1)
    cat("\n")
    
    report_proportion(x2, wt2)
    cat("\n")
    
    # report p-value of two sample t-test with weight
    # reformat data for chi-squared test
    id <- c(rep(1, length(x1)), rep(2, length(x2)))
    x <- c(x1, x2)
    wt <- c(wt1, wt2)
    
    tt <- weights::wtd.chi.sq(id, x, weight = wt)
    print(tt)
}

# example
x1 <- nhanes %>% filter(screened == 1) %>% pull(gender)
wt1 <- nhanes %>% filter(screened == 1) %>% pull(weight_interview)
x2 <- nhanes %>% filter(screened == 0) %>% pull(gender)
wt2 <- nhanes %>% filter(screened == 0) %>% pull(weight_interview)
compare_two_proportion(x1 == "F", x2 == "F", wt1, wt2)
```


## Report odds ratio from logistic regression

We expect to report odds ratio in a table like:

<table style="width:790px">
    <thead>
        <tr style="border-bottom:2px gray solid; color:black">
            <th align="left">Variable</th>
            <th colspan="2"; align="center">Overall population</th>
            <th colspan="2"; align="center">Screening recommended</th>
        </tr>
    </thead>
    <tbody>
        <tr style="border-bottom:1px gray solid; color:black"; align="left">
            <td> </td>
            <td>Odds ratio (95% CI)</td>
            <td>p value</td>
            <td>Odds ratio (95% CI)</td>
            <td>p value</td>
        </tr>
        <tr style="border-bottom:1px gray solid; color:gray"; align="left">
            <td>Age (year)</td>
            <td>1.03 (1.02-1.04)</td>
            <td>< 0.00001</td>
            <td>1.02 (1.01-1.04)</td>
            <td>0.0018</td>
        </tr>
        <tr style="border-bottom:1px gray solid; color:gray"; align="left">
            <td>Female</td>
            <td>1.24 (0.93-1.65)</td>
            <td>0.14</td>
            <td>1.09 (0.72-1.65)</td>
            <td>0.69</td>
        </tr>
    </tbody>
</table>
<br>


For a survey, the logistic regression must be weighted using `survey::svydesign` and and `survey::svyglm`. The `weights` in `glm` is **NOT** sampling weight, but the importance in error (or cost function in machine learning algorithms) calculation.


```{r}
get_odd_ratio <- function(lgr, n_decimal = 2){
    # To extract odds ratio with 95% confidence interval and p-value from a  
    # logistic regression model
    #
    # Arguments:
    # - lgr: logistic regression model
    # - n_decimal: interger, number of decimal to keep
    # 
    # Return: data frame containing odds ratio of each predictor
    
    estimate <- summary(lgr)$coefficients %>%
        data.frame() %>%
        # delete intercept
        .[-1, ] %>%
        mutate(odd_ratio = exp(Estimate) %>% round(2),
               CI_low = exp(Estimate - 1.96 * Std..Error) %>% round(2),
               CI_high = exp(Estimate + 1.96 * Std..Error)  %>% round(2),
               p_value = Pr...t..) %>%
        mutate(`Odds Ratio (95% CI)` = paste0(odd_ratio, 
                                              " (", 
                                              CI_low, 
                                              "-", 
                                              CI_high, 
                                              ")")) %>%
        select(`Odds Ratio (95% CI)`, p_value)
    
    return(estimate)
}

# example: survey based model
lgr_all <- svyglm(screened ~ age + gender + BMI + insured + hypertension + 
                      hyperlipidemia + physical_inactive + family_diabetes,
                  design = svydesign(ids = ~ 1, 
                                     data = nhanes, 
                                     weights = ~nhanes$weight_mec),
                  family = quasibinomial)  # to avoid warning with binomial
get_odd_ratio(lgr_all)
```



# Project 2: tbd
