---
title: "Geocoding Mailing Addresses to Census Blocks"
output: 
  html_notebook:
    number_sections: true
    toc: true
---

# Quick start

The best approach to geocode addresses to block geoids is to first geocode addresses to (lon, lat) coordinate using Arcgis and then geocode the coordinates into to block geoid using Census Geocoder. The approach combies  the Arcgis's ability to parsing addresses and Census Geocoder's capacity of linking census geography.

## Working example

In this example, we geocode 1000 addresses to (lon, lat) coordinates and then to block geoids.

```{r}
source("geocoding_utilities.R")
# sample address data, 1000 rows
data_addr <- fread("sample_addresses.csv")
   #      id                                        address
   # 1:    1  201 E GREENE ST, Milledgeville, GA, 31061, US
   # 2:    2  274 BETHEL FREEWILL RD, Baxley, GA, 31513, US
   # 3:    3   1217 BRECKINRIDGE TRL, Winder, GA, 30680, US

# from addresses to (lon, lat), create directory "sample_lon_lat"
arcgis_geocode(data = data_addr, 
               address = "address",
               save_dir = "sample_lon_lat/")

# open another R session and geocode above (lon, lat) to block geoids. Create a
# new directory "sample_block_geoid". n_batch is known from arcgis_geocode()
lon_lat_to_geoid_concurrent(batch_dir_from = "sample_lon_lat/",
                            batch_dir_to = "sample_block_geoid/",
                            n_batch = 15)
```


## Geocoding utility functions

These functions are saved in a file called `geocoding_utilities.R`.


```{r}
# Step 1: from addresses to (lon, lat) coordinates ===========================
# 
# These functions use `tidygeocoder` package to call Arcgis API. It takes about 
# 1 mins to geocode 100 addresses.

arcgis_geocode_mini <- function(data, address, 
                                save_as = "tmp.csv"){
  # To geocode a small number of address to (lon, lat) coordinates. The 
  # coordinates are appended to the original data as two new columns and saved 
  # as a csv file.
  #
  # Arguments:
  # - data: data frame containing an address column. The addresses are in the 
  #   format like "123 Hope St,Boston,MA,05678,US". Missing values are allowed
  #   in places other than street addresses. Works best when the data frame has
  #   about 70 addresses. Much slower with larger number of addresses.
  # - address: string, column name of the address column.
  # - save_as: string, file path to save the results.
  #
  # Return: data frame with geocoded coordinates
  
  # slow process. Monitor the time.
  t1 <- Sys.time()
  
  res <- data %>%
    as.data.frame() %>%
    tidygeocoder::geocode(
      address = address,
      method = "arcgis",
      lat = "lat",
      long = "lon")
  
  t2 <- Sys.time()
  print(t2 - t1)
  
  data.table::fwrite(res, file = save_as)
  
  return(invisible(res))
}


arcgis_geocode <- function(data, 
                           address,
                           mini_batch = 70,
                           save_dir = "tmp/",
                           start_from = 1){
  # To geocode a large number of address to (lon, lat) coordinates. The 
  # coordinates are appended to the original data as two new columns and saved 
  # as multiple csv files in a directory.
  #
  # Arguments:
  # - data: data frame containing an address column. The addresses are in the 
  #   format like "123 Hope St,Boston,MA,05678,US". Missing values are allowed
  #   in places other than street addresses. 
  # - address: string, column name of the address column.
  # - mini_batch: integer, number of rows of the data frame fed to function
  #   arcgis_geocode_mini().
  # - save_dir: string, directory to save the geocoded coordinates.
  # - start_from: which batch to start from. It allows continuing geocoding 
  #   from last batch in case the process is interrupted.
  #
  # Return: NULL
  
  n_minibatch <- nrow(data) %/% mini_batch
  remainder <- nrow(data) %% mini_batch
  
  for (i in start_from:n_minibatch){
    print(paste0(i, "/", n_minibatch))
    file_name <- paste0(save_dir, "batch_", i, ".csv")
    arcgis_geocode_mini(data[(mini_batch * (i - 1) + 1):(mini_batch * i), ], 
                        address,
                        save_as = file_name)
  }
  
  if (remainder > 0){
    arcgis_geocode_mini(
      data[(mini_batch * n_minibatch + 1):(mini_batch * n_minibatch + remainder), ],
      address,
      save_as = paste0(save_dir, "batch_", n_minibatch + 1, ".csv")
    )
  }
}



# Step 2: from (lon, lat) to block geoid ======================================

# At this moment we geocode census 2010 geoids. Pay attention to the new 2020 
# geoids when they are available.
# 
# This part uses `censusxy` package.

get_block_geoid_single <- function(lon, lat, vintage = "Census2010_Current"){
  # To query the block geoid of a single (lon, lat) coordinate
  # 
  # Arguments:
  # - lon, lat: numeric
  # - vintage: same as vintage in censusxy::cxy_geography
  #
  # Return: string, census block geoid like "10000US440116543211002"
  
  if (is.na(lon) || is.na(lat)){
    return(NA)
  }
  
  geoid <- censusxy::cxy_geography(lon, lat, vintage = vintage) %>%
    .[1, "Census.Blocks.GEOID"] 
  
  if (length(geoid) == 0){
    return(NA)
  }
  
  paste0("10000US", geoid)
}

# # example
# block_2020 = get_block_geoid_single(lon = -110.23324, lat = 38.63593, "Census2020_Current")
# block_2010 = get_block_geoid_single(lon = -110.23324, lat = 38.63593)

get_block_geoid_mini <- function(data, lon = "lon", lat = "lat", 
                                 vintage = "Census2010_Current",
                                 save_as = "tmp.csv"){
  # To geocode a small number of (lon, lat) coordinates into block geoids. The 
  # geoids are appended to the original data as a new column and saved 
  # as a csv file.
  #
  # Arguments:
  # - data: data frame containing two columns for longitude and latitude. 
  #   Works best when the data frame has about 70 addresses. Much slower with
  #   with larger number of coordinates.
  # - lon, lat: string, column name of longitude ant latitude.
  # - vintage: same as vintage in censusxy::cxy_geography
  # - save_as: string, file path to save the results.
  #
  # Return: data frame with encoded block geoids
  
  t1 <- Sys.time()
  res = c()
  for (i in 1:nrow(data)){
    geoid <- get_block_geoid_single(data[[lon]][i], data[[lat]][i], vintage)
    res <- c(res, geoid)
  }
  data$block_geoid <- res
  
  t2 <- Sys.time()
  print(t2 - t1)
  
  fwrite(data, file = save_as)
  return(invisible(data))
}


# bbb <- get_block_geoid_mini(lon_lat[1:10,])


census_geocode_lon_lat <- function(data, 
                                   lon = "lon", 
                                   lat = "lat",
                                   vintage = "Census2010_Current",
                                   mini_batch = 70,
                                   saved_dir = "tmp/",
                                   start_from = 1,
                                   file_name = NULL){
  # To geocode a large number of (lon, lat) coordinates to block geoids. The 
  # block geoids are appended to the original data as a new column and saved 
  # as multiple csv files in a directory.
  #
  # Arguments:
  # - data: data frame containing longitude and latitude columns. 
  # - lon, lat: string, column name of lon and lat.
  # - vintage: same as vintage in censusxy::cxy_geography
  # - mini_batch: integer, number of rows of the data frame fed to function
  #   arcgis_geocode_mini().
  # - save_dir: string, directory to save the geoids.
  # - start_from: which batch to start from. It allows continuing geocoding 
  #   from last batch in case the process is interrupted.
  #
  # Return: NULL
  
  n_minibatch <- nrow(data) %/% mini_batch
  remainder <- nrow(data) %% mini_batch
  
  for (i in start_from:n_minibatch){
    print(paste0(i, "/", n_minibatch))
    
    if (is.null(file_name)){
      file_name <- paste0(saved_dir, "batch_", i, ".csv")
    }
    
    get_block_geoid_mini(data[(mini_batch * (i - 1) + 1):(mini_batch * i), ], 
                        lon, lat, vintage,
                        save_as = file_name)
  }
  
  if (remainder > 0){
    get_block_geoid_mini(
      data[(mini_batch * n_minibatch + 1):(mini_batch * n_minibatch + remainder), ],
      lon, lat, vintage,
      save_as = paste0(saved_dir, "batch_", n_minibatch + 1, ".csv")
    )
  }
}

lon_lat_to_geoid_concurrent <- function(batch_dir_from, batch_dir_to, 
                                        n_batch, start_from = 1){
  # geocode (lon, lat) to block geoid while (lon, lat) batch files are being
  # created by arcgis_geocode(). 
  #
  # Arguments:
  # - batch_dir_from: directory where geocoded (lon, lat) files are saved
  # - batch_dir_to: directory where geocided block geoid files are saved
  # - n_batch: number of file expected in batch_dir_from
  # - start_from: which batch to start from 
  
  for (i in start_from:n_batch){
    print(paste0(i, " / ", n_batch))
    fpath <- paste0(batch_dir_from, "batch_", i, ".csv")
    while(!file.exists(fpath)){
      message("Waiting for new (lon, lat) file: ", fpath)
      Sys.sleep(30)
    }
    
    lon_lat <- fread(fpath)
    
    fname <- paste0(batch_dir_to, "batch_", i, ".csv")
    census_geocode_lon_lat(lon_lat, 
                           saved_dir = batch_dir_to,
                           file_name = fname)
  }
}


# Helper functions =============================================================

read_batch_geocoding <- function(data_dir){
  # To read all csv files in directory into a data.table
  fnames <- paste0(data_dir, list.files(data_dir))
  res <- data.table::fread(fnames[1])
  for (i in 2:length(fnames)){
    if (i %% 100 == 0){
      print(i)
    }
    tmp <- fread(fnames[i])
    res <- rbindlist(list(res, tmp))
  }
  
  return(res)
}
```


# Census geocoding fundamentals

## Benchmark and vintage


# More details

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Geocoding address to census block
**What & Why**: Very often a location is given as a mailing address, from which we know the city, state, and zip code. More refined information, however, are associated with the street numbers. For example, the census data at census tract, block group, and block levels. Census geocoding is a technique that map a street address to a census block, in addition to simply its (lon, lat) coordinate.

**How**: [Census Geocoder](https://geocoding.geo.census.gov/geocoder/geographies/addressbatch?form) provides unlimited free service. ArcGIS has very reliable coordinate [geociding](https://developers.arcgis.com/documentation/mapping-apis-and-services/search/services/geocoding-service/), but extra effort is needed to map the address to census geographies. Other geocoding providers can be found [here](https://dlab.berkeley.edu/blog/address-geocoding-options-uc-berkeley-community).

## Census Geocoder
There are two ways to use Census Geocoder: the online service or a programming language.

### Online service
To [census geocode a batch of addresses](https://geocoding.geo.census.gov/geocoder/geographies/addressbatch?form), select `FIND GEOGRAPHIES USING ...` option and upload a csv file containing the addresses. Each line follows the format: id, street address, city, state, zipcode. Missing values are allowed but must seperated with comma, like the third line in the sample below.

```
1,4600 Silver Hill Road,Washington,DC,20233
2,400 15th St SE,Washington,DC,20003
3,72 Hope St,,RI,
```

After uploading the csv file, click on `Get Results` to download the results in a csv file. It may take a very long time, depending on how many addresses in the file. As a rough estimate, it takes about 10 min for 1000 addresses. So be patient.


### API????
The address batch API is 
```
https://geocoding.geo.census.gov/geocoder/geographies/addressbatch?form
```
To make a API call, replace `?form` with search parameters.


### R censusxy package
The [censusxy](https://github.com/slu-openGIS/censusxy/) package is dedicated to Census Geocoder. Like tidycoder, there are many random no-matches when used for batch geocoding (a problem with the api????????). The censusxy package, however, allows parallel query so it is significantly faster. With the fast speed, we can afford repeated geocoding until there is no improvement, as shown in the  functions below. 

#### Batch geocoding with addresses

**Benchmark and vintage**: 
More details are [here](https://www.census.gov/programs-surveys/geography/technical-documentation/complete-technical-documentation/census-geocoder.html).

- Benchmark is the time period a snapshot of data is created. Run `cxy_benchmarks()` to check available benchmarks. 
- Vintage of geography is the census or survey the data relates to. Run `cxy_vintages("a_benchmark_name")` to check for available vintages of this benchmark.
- Under "Public_AR_Current" benchmark, we have the following vintages of interests:
- Census2010_Current for Census2010 Vintage
- ACS2019_Current for ACS2019 Vintage
- Census2020_Current for Census2020 Vintage. The block id is **completely different** from Census2010.

```{r eval=FALSE}
library(censusxy)
library(dplyr)


census_geocode_mini <- function(data, 
                                id = NULL,
                                street,
                                city = NULL,
                                state = NULL,
                                zip = NULL,
                                return = "geographies",
                                vintage = "Census2010_Current",
                                n_repeat = 4, 
                                n_jobs = 6,
                                save_as = "tmp.csv"){
  # To geocode the geograhies of a small number of address. The Census Geocoder API is very slow when 
  # submit a large batch. 
  #
  # Argument:
  # - data: data.frame containing columns for addresses, with about 70 rows
  # - n_repeat: repeated time without improvement before stopping query
  # - n_job: cores for parallel query
  # - save_as: save the geocided data into a csv file.
  # - other arguments: check cxy_geocode function
  
  t1 <- Sys.time()
  N0 <- nrow(data)
  
  # create a placeholder for geocoded data
  df0 <- data[FALSE, ]
  df1 <- structure(list(cxy_lon = numeric(0), cxy_lat = numeric(0), 
                        cxy_state_id = integer(0), cxy_county_id = integer(0), cxy_tract_id = integer(0), 
                        cxy_block_id = integer(0)), row.names = integer(0), class = "data.frame")
  matched <- cbind(df0, df1)
  
  # repeated geocoding until not impovement after n_repeat tries
  keep_downloading = TRUE
  i_round <- 0
  i_repeat <- 0
  while(keep_downloading){
    i_round <- i_round + 1
    print(i_round)
    downloaded <- cxy_geocode(data, id, street, city, state, zip,
                              output = "simple", 
                              return = return, 
                              vintage = vintage, 
                              parallel = n_jobs)
    
    matched <- rbind(matched, downloaded[!is.na(downloaded$cxy_lon), ])
    
    print(paste0("---- unmatched: ", N0 - nrow(matched), " ----"))
    
    if (nrow(matched) < N0){
      no_match <- sum(is.na(downloaded$cxy_lon))
      
      if (no_match == nrow(data)){
        i_repeat <- i_repeat + 1
        print(paste0("i_repeat = ", i_repeat))
        if (i_repeat >= n_repeat){
          keep_downloading <- FALSE
        }
      } else {
        i_repeat <- 0
      }
      # new data
      data <- downloaded[is.na(downloaded$cxy_lon), names(df0)]
    } else {
      keep_downloading = FALSE
    }
  }
  
  t2 <- Sys.time()
  print(t2 - t1)
  
  message(N0 - nrow(matched), " addresses have no match")
  
  if (nrow(matched) == N0){
    res <- matched
  } else {
    res <- rbind(matched, downloaded)
  }
  
  res <- res %>%
    rename(lon = cxy_lon,
           lat = cxy_lat) %>%
    mutate(block_geoid = if_else(
      is.na(lon), 
      NA_character_, 
      paste0("10000US", cxy_state_id, cxy_county_id, cxy_tract_id, cxy_block_id)
    ))
  
  data.table::fwrite(res, file = save_as)
  return(res)
}

# # # example
# aaa = census_geocode_mini(stl_homicides[1:70, ],
#                           street = "street_address",
#                           city = "city",
#                           state = "state")


census_geocode <- function(data,
                           id = NULL,
                           street,
                           city = NULL,
                           state = NULL,
                           zip = NULL,
                           return = "geographies",
                           vintage = "Census2010_Current",
                           n_repeat = 4, 
                           n_jobs =6,
                           mini_batch = 70,
                           saved_dir = "tmp/",
                           start_from = 1){
  # To geocode a large number of addresses. The dataframe is broken into n_minibatch small dataframe 
  # and queried and saved as csv files separated. 
  #
  # Argument:
  # - mini_batch: number of row of a mini_batch
  # - saved_dir: directory to save the geocoded data
  # - start_from: integer, the first batch to start from. Used when connection
  #   broken.
  # - other arguments: see census_geocode_mini function
  
  n_minibatch <- nrow(data) %/% mini_batch
  remainder <- nrow(data) %% mini_batch
  
  for (i in start_from:n_minibatch){
    print(paste0(i, "/", n_minibatch))
    file_name <- paste0(saved_dir, "batch_", i, ".csv")
    census_geocode_mini(data[(mini_batch * (i - 1) + 1):(mini_batch * i), ], 
                        id, street, city, state, zip,
                        return = return, 
                        vintage = vintage, 
                        n_repeat = n_repeat,
                        n_jobs = n_jobs,
                        save_as = file_name)
  }
  
  if (remainder > 0){
    census_geocode_mini(data[(mini_batch * n_minibatch + 1):(mini_batch * n_minibatch + remainder), ],
                        id, street, city, state, zip,
                        return = return, 
                        vintage = vintage, 
                        n_repeat = n_repeat,
                        n_jobs = n_jobs,
                        save_as = paste0(saved_dir, "batch_", n_minibatch + 1, ".csv"))
  }
}

# # example
# census_geocode(stl_homicides[1:100,],
#                street = "street_address",
#                city = "city",
#                state = "state",
#                saved_dir = "tmp/")

```

#### batch geocoding with (lon, lat) coordinate
The `cxy_geography()` function in censusxy package allows geocoding of geographies from a single (lon, lat) coordinate.  
```{r}
get_block_geoid_single <- function(lon, lat, vintage = "Census2010_Current"){
  # To query the block geoid of a single coordinate
  if (is.na(lon) || is.na(lat)){
    return(NA)
  }
  geoid <- cxy_geography(lon, lat, vintage = vintage) %>%
    .[1, "Census.Blocks.GEOID"] 
  
  if (length(geoid) == 0){
    return(NA)
  }
  
  paste0("10000US", geoid)
}

# # example
# block_2020 = get_block_geoid_single(lon = -110.23324, lat = 38.63593, "Census2020_Current")
# block_2010 = get_block_geoid_single(lon = -110.23324, lat = 38.63593)

get_block_geoid <- function(data, lon = "lon", lat = "lat", 
                            vintage = "Census2010_Current"){
  # the data has three column, id, lon, and lat
  res = data.frame()
  for (i in 1:nrow(data)){
    print(i)
    print(Sys.time())
    geoid <- get_block_geoid_single(data[[lon]][i], data[[lat]][i], vintage)
    res <- c(res, geoid)
  }
  data$block_geoid <- res
  return(data)
}


# example
df <- structure(list(addr_id = 1:6, 
                     lat = c(29.7459090034226, 29.7293539902798, 
                             29.7191180101219, 29.725406993984, 
                             29.7475459891127, 29.7864580106858), 
                     lon = c(-95.3695059732392, -95.3862349946919, 
                             -95.3815090262245, -95.3689899832797, 
                             -95.3768780247201, -95.3634529818611)), 
                row.names = c(NA,-6L), 
                class = "data.frame")

bbb = get_block_geoid(df, lon = "lon", lat = "lat")
ccc = get_block_geoid(df, lon = "lon", lat = "lat", "Census2020_Current")
```


## R tidygeocoder package
The tidygeocoder package works with several geocoding providers. 

### Census Geocoder
This package is not as good as the censusxy package. Below shows the procedure of geocoding geographies but we are not going to use it.

```{r}
library(tidygeocoder)
library(magrittr)

df <- structure(list(addr_id = 1:6,
                     address = c(
                       "2222 Austin St, houston, TX, 77002",
                       "5001 Main St, houston, TX, 77002",
                       "1 Hermann Museum Circle Dr, houston, TX, 77004",
                       "2612 Wheeler St, houston, TX, 77004",
                       "2450 Louisiana St Ste 915, houston, TX, 77006",
                       "822 Booth St, houston, TX, 77009"
                     )),
                row.names = c(NA, -6L),
                class = "data.frame")

census_block <- df %>%
  geocode(address,
          method = "census",
          return_type = "geographies",
          full_results = TRUE)

knitr::kable(census_block)
```

The code looks straightforward, but the actual query process is more complicated than it appears, as different run returns different results, with no-match from time to time. If you do a batch query of 1000 addresses, it is no surprise if the return has one third no-matches. You will have to separate those not-matched addresses and query again. To make it even more complicated, the query time does not linearly increase with number of addresses. From my experience, batch size of 100 is very efficient and batch size of 4 tends to give less no-matches. 

**Step 1**: So here is the first round of geocoding if you have large number of addresses. Remember to save each query into a csv file for later.

```{r eval=FALSE}
# assuming address_df is a data.frame with many rows and the column address is 
# for one-line addresses as in mail address, and also has a address_id column
# to track the query results.

for (i in 1:100){
  batch_size = 100
  print(batch_size)
  file_name <- paste0("batch_geocode/batch_", i, ".csv")
  
  t1 <- Sys.time()
  tmp <- address_df[(batch_size*(i - 1) + 1):(batch_size*i), ] %>%
    geocode(address,
            method = "census",
            return_type = "geographies",
            full_results = TRUE,
            verbose = FALSE)
  t2 <- Sys.time()
  print(t2 - t1)
  
  fwrite(tmp, file = file_name)
}
```

**Step 2**: Separate matched and not-matched address

```{r eval=FALSE}
library(data.table)
library(magrittr)

df_matched <- fread("batch_geocode/batch_1.csv") %>%
  .[!is.na(match_type)]
id_not_matched <- fread("batch_geocode/batch_1.csv") %>%
  .[!is.na(match_type), address_id]

# i starts from 2
for (i in 2:100){
  file_name <- paste0("batch_geocode/batch_", i, ".csv")
  tmp <- fread(file_name)
  
  df_matched <- rbindlist(list(df_matched, tmp[!is.na(match_type)]))
  id_not_matched <- append(id_not_matched, tmp[is.na(match_type), address_id])
}

fwrite("matched_results/matched_round_1.csv")
```

**Step 3**: Repeat step 1 and 2 for unmatched addresses until no improvement.

### ArcGIS and Google
We will use ArccGIS to extract (lon, lat) only. The coordinates are then fed to censusxy to download block geoid.

```{r}
lon_lat <- df %>%
  geocode(address,
          method = "arcgis",
          long = "lon",
          lat = "lat")

lon_lat <- df %>%
  geocode(address,
          method = "google",
          long = "lon",
          lat = "lat")
```


## Guessing races from names